Modelling palaeoecological time series using generalised additive models

### 2018-04-28 manuscript.Rmd
updated manuscript

### Data

intro on the two data sets goes here.

@@

The second example time series is a 6,000 year record of alkenone unsaturation, \uk, from Braya-sø, a meromictic lake in  West Greenland [@DAndrea2011-oa]. Alkenones are long-chained unsaturated organic compounds that are produced by a small number of planktonic organisms known as haptophytes. The \uk{} unsaturation index is

\begin{equation*}
\uk{} = \frac{[C_{37:2}] - [C_{37:4}]}{[C_{37:2}] + [C_{37:3}]+ [C_{37:4}]}
\end{equation*}

where $[C_{37:x}]$ is the concentration of the alkenone with 37 carbon atoms and $x$ double carbon bonds. The relative abundance of these alkenones is known to vary with changes in water temperature, and as a result \uk{} is used as a proxy for lake- and sea-surface temperatures. For further details on the Braya-sø \uk{} record and age model see @DAndrea2011-oa. Here I use the 3,000 year \uk{} record from the PAGES 2K database [@PAGES_2k_Consortium2013-fm]. The data are presented in Figure \ref{fig:data-figure}b.

@@

### Regression models for palaeoenvironmental time series

A linear model for a trend in a series of $T$ observations $y_t$ at observation times $x_t$ with $t = 1, 2, \ldots, T$ is

\begin{equation} \label{eq:linear-model}
y_t = \beta_0 + \beta_1 x_t + \varepsilon_t,
\end{equation}

where $\beta_0$ is a constant term, the model *intercept*, representing the expected value of $y_t$ where $x_t$ is 0. $\beta_1$ is the *slope* of the best fit line through the data; it measures the rate of change in $y$ for a unit increase in $x$. The unknowns, the $\beta_j$ are commonly estimated using least squares by minimising the sum of squared errors, $\sum_t \varepsilon_t^2$. If we want to ask if the estimated trend $\beta_1$ is statistically significant, a process called *inference*, we must make further assumptions about the data (conditional upon the fitted model) or the model errors (residuals); $\varepsilon_t \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2)$. This notation indicates that the residuals $\varepsilon_t$ are *independent* and *identically distributed* Gaussian random variables with mean equal to $0$ and constant variance $\sigma^2$. In the time series setting, the assumption of independence of model residuals is often violated.

The linear model described above is quite restrictive in terms of the types of trend it can fit; essentially linear increasing or decreasing trends, or trivially, a null trend of no change. This model can be extended to allow for non-linear trends, most notably via making $y_t$ depend on polynomials of $x_t$, for example

\begin{align} \label{eq:polynomial-model}
y_t &= \beta_0 + \beta_1 x_t + \beta_2 x_t^2 + \cdots + \beta_P x_t^P + \varepsilon_t \\
    &= \beta_0 + \sum_{p = 1}^P \beta_p x_t^p  + \varepsilon_t \nonumber %\\
    %&= \boldsymbol{\beta} \mathbf{X} + \boldsymbol{\varepsilon}
\end{align}

where polynomials of $x_t$ up to order $P$ are used. This model allows for more complex fitted trends but it remains a fully parametric model and suffers from several problems, especially the behaviour of the fitted polynomial at the start and end of the observed series.

@@

Linear models employing a range of polynomials of the covariate Year for the Small Water data set are shown in Figure \ref{fig:polynomial-example-plot}. The low-order models ($P \in \{1, 3\}$) result in very poor fit to the data. The model with $P = 5$ does a reasonable job of capturing the gross pattern in the data, but it fails to adapt quickly enough to the decrease in δ^15^N that begins ~1940 CE, and the resulting trend is quite biased as a result. The $P = 10$th-order polynomial model is well able to capture this period of rapid change, but it does so at the expense of increased complexity in the estimated trend prior to ~1940. Additionally, this model ($P = 10$) has undesirable behaviour at the ends of the series, significantly overfitting the data, a commonly observed in polynomial models such as these. Finally, the choice of what order of polynomial to fit is an additional choice left for the analyst to specify; choosing the optimal $p$ is not a trivial task when the data are a time series and residual autocorrelation is likely present.

@@

Can we do better than these polynomial fits? In the remainder , I hope to demonstrate that the answer to that question is emphatically "yes". Below I describe a coherent and consistent approach to modelling palaeoenvironmental time series using generalised additive models (GAMs), which builds upon the linear regression framework.

@@

GAMs extend the generalised linear model (GLM) --- of which the linear model described above is a special case --- where some or all of the parametric terms, the $\beta_p$, are replace by smooth functions $f_j$ of the covariates $x_j$. The GAM version of \eqref{eq:linear-model} is

@@

where $\mu_t$ is the expected value of the random variable $Y_t$ ($\mu_t \equiv \mathbb{E}(Y_t)$) of which we have observations $y_t$. $g$ is the link function, an invertible, monotonic function, such as the natural logarithm, and $g^{-1}$ its inverse. The link function maps values from the linear predictor scale on to the scale of the response. For example, count data are strictly non-negative integer values and are commonly modelled as a Poisson GLM/GAM using the natural log link function. On the log scale, the response can take any real value between $-\infty$ and $+\infty$, and it is on this scale that model fitting actually occurs. However we need to map these unbounded back on to the non-negative response scale the. The inverse of the log link function is used, the exponential function, maps values to the interval 0--$\infty$.

In \eqref{eq:gam}, we further assume that the observations are drawn from a member of the exponential family of distributions --- such as the Poisson for count data, the binomial for presence/absence or counts from a total --- with expected value $\mu_t$ and possibly some additional parameters $\Theta$ ($y_t \sim \text{EF}(\mu_t, \Theta)$). Additionally, many software implementations of the above models now allow for the fitting of distributions that are not within the exponential family but which can be fitted using an algorithm superficially similar to the one used to fit GAMs to members of the exponential family. Common examples of such extended families include the negative binomial distribution (for overdispersed counts) and the beta distribution (for true proportions or other interval-bounded data).

@@

We want the fitted trends for the Small Water δ^15^N and Braya-sø \uk{} time series to be wiggly functions, whose shape is informed by the data themselves. These wiggly functions are smooths, which we denote by $f(x_t)$. We need to represent the smooths in a way that (\ref{eq:gam}) is a linear model, which is achieved by using a *basis*. A basis is a set of functions that collectively span a space of functions that, we hope, contains the true $f(x_t)$ or a close approximation to it. The functions in the basis are known as *basis functions*, and arise from a *basis expansion* of a covariate. Writing $b_j(x_t)$ as the $j$th basis function of $x_t$, the smooth $f(x_t)$ can be represented as a weighted sum of basis functions

$$
f(x_t) = \sum_{j = 1}^{k} b_j(x_t) \beta_j
$$

@@

The polynomial model we encountered earlier is an example of a statistical model that uses a basis expansion. For the cubic polynomial ($P = 3$) fit shown in Figure \ref{fig:polynomial-example-plot} there are in fact 4 basis functions: $b_1(x_t) = x_t^0 = 1$, $b_2(x_t) = x_t$, $b_3(x_t) = x^2_t$, $b_4(x_t) = x_t^3$. Note that here $b_1(x_t)$ is constant and is in fact the model intercept, $\beta_0$, in (\ref{eq:polynomial-model}), and the weights are the estimated coefficients in the model, the $\beta_j$.

As we have already seen, polynomial basis expansions do not necessarily lead to well-fitting models unless the true function $f$ is itself a polynomial. One of the primary criticisms is that polynomial basis functions are global; the value of $f$ at time point $x_t$ affects the value of $f$ at time point $x_{t+n}$ even if the two time points are at opposite ends of the series. There are many other bases we could use; here I discuss one such set of bases, that of splines.

@@

A CRS can be parameterized in a number of different ways. One requires a knot at each unique data value in $x_t$, which is computationally inefficient. One approach to specifying a CRS basis is to parameterize in terms of the value of the spline at the knots. Typically in this parametrization there are many fewer knots than in alternative parameterizations, and these knots are either distributed evenly over the range of $x_t$ or at the quantiles of $x_t$. Placing knots at the quantiles of $x_t$ has the effect of placing a greater number of knots where the data is densest.

@@

Cubic regression splines, as well as many other types of spline, require the analyst to choose the number and location of the knots that parametrise the basis. Thin plate regression splines (TPRS) remove this element of subjectivity when fitting GAMs. Thin plate splines were introduced by @Duchon1977-jr and, as well as solving the knot selection problem, have several additional attractive properties in terms of optimality and their ability to estimate a smooth function of two or more variables, leading to the inclusion of smooth interactions between covariates. However, thin plate splines have one key disadvantage over CRS; thin plate splines have as many unknown parameters as there are unique combinations of covariate values in a data set. It is unlikely in the extreme that any real data problem would require functions of such complexity that they require so many basis functions. It is much more likely that the true functions that we attempt to estimate are far simpler than the set of functions able to be represented by 1 basis function per unique data value. From a practical point of view, it is highly inefficient to carry around all these basis functions whilst model fitting, and the available computational resources would become quickly exhausted for large time series with many observations.

@@

The truncation suggested by @Wood2003-qy is not without cost; the eigen-decomposition and related steps can be relatively costly to set up for large data sets. For data sets of similar size to the two examples used here, the additional computational effort required to set up the thin plate regression spline basis over say the cubic regression spline basis will not be noticeable. For highly resolved data sets containing more than 500--1000 observations the truncation may be too costly computationally. In such instances, little is lost by moving to the cubic regression spline basis, with the same number of knots as the rank of the desired thin plate spline, either distributed evenly throughout the data or at the quantiles of the data, with the benefit of considerably reduced set up time for the basis.

With the two regression spline bases described above, to fit a GAM the analyst is generally only required to the specify the size (rank, number of knots) of the basis expansion required to represent or closely approximate the true function $f$. With practice and some knowledge of the system from which the observations arise, it can be relatively easy to put an upper limit on the expected complexity of the true trend in the data. Additionally, the number available data points places a constraint on the upper limit of the size of basis expansion that can be used.

In practice, the size of the basis is an upper limit on the expected complexity of the trend, and a simple test is available to check if the basis used was sufficiently large [@Pya2016-rk]. This test is available via the `gam.check()` function in **mgcv** for example, and essentially looks at whether there is any additional nonlinearity or structure in the residuals that can be explained by $x_t$. Should a smooth term in the fitted model fail this test the model can be refitted using a larger basis expansion, say by doubling the value of `k` used to fit the original. Note also that a smooth might fail this test whilst using fewer effective degrees of freedom than the maximum possible for the dimension of basis used. This may happen when the true function lies at the upper limit of the set of functions encompassed by the size of basis used. Additionally, a basis of size $2k$ encompasses a larger space of functions of a given complexity than a basis of size $a$ [@Wood2017-qi]; increasing the basis dimension used to fit the model may unlock this additional function space resulting in a better fitting model whilst using a similar number of effective degrees of freedom.

@@

Having identified regression splines as a useful way to represent $f$, we next need a way to decide how wiggly the fitted trend should be. A backwards elimination approach to sequentially remove knots or basis functions might seem appropriate, however such an approach would likely fail as the resulting sequence of models would not strictly nested, precluding many forms of statistical comparison. Alternatively, we could keep the basis dimension at a fixed size but gaurd against fitting very complex models through the use of a wiggliness penalty.

@@

$$\mathcal{L}_p(\boldsymbol{\beta}) = \mathcal{L}(\boldsymbol{\beta}) - \frac{1}{2} \lambda\boldsymbol{\beta}^{\mathsf{T}}\mathbf{S}\boldsymbol{\beta}$$

@@

To fully automate smoothness selection for $f(x_t)$ we need to estimate $\lambda$. There are two main ways that $\lambda$ can be automatically chosen during model fitting. The first way is to choose $\lambda$ such that it minimises the prediction error of the model. This can be achieved by choosing $\lambda$ to minimise Akaike's information criterion (AIC) or via cross-validation (CV) or generalized cross-validation (GCV). GCV avoids the computational overhead inherent to CV of having to repeatedly refit the model with one or more observations left out as a test set. Minimising the GCV score will, with a sufficiently large data set, find a model with the minimal prediction error. The second approach is to treat the smooth as a random effect, in which $\lambda$ is now a variance parameter to be estimated using maximum likelihood (ML) or restricted maximum likelihood (REML).

@@

### Smoothing time series data

**Do I need this as a separate section?** Or just cover the points as they occur?

## Fitting GAMs

@@

Note that now I have relaxed the i.i.d. assumption and have introduced $\boldsymbol{\Lambda}$, a correlation matrix that is used to model autocorrelation in the residuals. The δ^15^N values are irregularly spaced in time and as a result a correlation structure that can handle the uneven spacing of the samples is needed. A continuous time first-order autoregressive process (CAR(1)) is a reasonable choice; it is the continuous-time equivalent of the first-order autoregressive process (AR(1)) and, simply stated, models the correlation between any two residuals as $\phi^h$, where $h$ is the amount of separation in time between the residuals. $h$ may be a real valued number in the CAR(1), which accommodates the irregular separation of samples. $\phi$ controls how quickly the correlation between any two residuals declines as a function of their separation in time. $\phi$ is an additional parameter that will be estimated during model fitting. The model in \ref{eq:small-gam} was fitted using the `gamm()` function [@Wood2004-zv] in the **mgcv** package [@Wood2017-qi] for R [@R-core-team].

@@

## convert to gtabl to align heights
g_small_fitted <- ggplotGrob(small_fitted) # convert to gtable
g_braya_fitted <- ggplotGrob(braya_fitted) # convert to gtable

height_take <- c(4,5)
small_heights <- g_small_fitted$heights[height_take]
braya_heights <- g_braya_fitted$heights[height_take]
max_heights <- unit.pmax(small_heights, braya_heights) # calculate maximum heights
g_small_fitted$heights[height_take] <- max_heights # assign max heightss to small gtable
g_braya_fitted$heights[height_take] <- max_heights

plot_grid(g_small_fitted, g_braya_fitted, ncol = 1, labels = "auto", align = "hv", axis = "lr")
```

The fitted trend is shown in Figure \ref{fig:plot-fitted-models}a, and well-captures the strong pattern in the data. The trend is statistically significant (estimated degrees of freedom = `r round(smallSumm[["s.table"]][1,2], 3)`; $F$ = `r round(smallSumm[["s.table"]][1,3], 3)`, approximate $p$ value = $\ll$ 0.0001). However further analysis of the fitted model is required to answer the other questions posed earlier about the timing of change and whether features in the trend can be distinguished from random noise.

@@

The \uk{} data present a more difficult data analysis challenge than the δ^15^N time series because of the much more complex variation present. Fitting the same model as the Small Water example, \ref{eq:small-gam}, to the \uk{} data resulted in the unsatisfactory fit shown as the very smooth line in Figure \ref{fig:plot-fitted-models}b (labelled GAMM (CAR(1))). Further problems were evident with this model fit --- the covariance matrix of the model was non-positive definite, a sure sign of problems with the fitted model. Refitting with a smaller basis dimension (*k* = 20) for the trend term resulted in a model with a positive-definite covariance matrix for the model variance-covariance terms, but the estimated value of of the CAR(1) parameter $\phi$ = `r round(brayaPhi[2], 2)` was exceedingly uncertain (95% confidence interval `r round(brayaPhi[1], 2)` -- `r round(brayaPhi[3], 2)`!)

@@

To investigate the first issue the GCV and REML scores for an increasing sequence of values of the smoothness penalty $\lambda$ were evaluated for the standard GAM (equation \ref{eq:gam}) fit to the \uk{} time series. The resulting profiles are shown in Figure \ref{fig:trace-smoothness-parameters}, with the selected value of the penalty shown by the vertical line. The GCV score profile suggests that the potential for under smoothing identified by @Reiss2009-fk is unlikely to apply here. Addressing the violation of the independence assumption is more difficult; the standard errors of the model coefficients are most likely anti-conservative, which presents a significant problem if we wish to determine which of the wiggles in the fitted \uk{} trend can be identified as real.

@@

While there is not much we can do within the GAM framework to model a series that contains both smooth trends and step-like responses, an adaptive smoother can help address problems where the time series consists of periods of very rapid change in the mean combined with periods of complacency or relatively little change. As suggested by their name, adaptive smoothers can adjust to changes in the wiggliness of the time series. This adaptive behaviour is achieved by making the smoothness parameter $\lambda itself depend smoothly on $x_t$; in other words, the adaptive smoother allows the wiggliness of the estimated trend to vary smoothly over time. This is achieved by using multiple penalties and smoothness parameters, which have the effect of letting the strength of the wiggliness vary smoothly over time. Whilst this does allow the estimated trend to adapt to periods of rapid change in the response, adaptive smoothers make significant demands on the data; if we used $m$ smoothness penalties to allow the wiggliness to vary over a time series, it would be like estimating $m$ separate smooths from chunks of the original series each of length $n/m$. In a practical sense, this limits the use of adaptive splines in palaeoecology to proxies that are readily enumerated, such as the biogeochemical proxies used in the two example data sets.

@@

Figure \ref{fig:small-scam-fit-plots}a shows the estimated dates (in Years CE) for `r nrow(swAge)` levels in the Small Water core dated using ^210^Pb. The vertical bars show the estimated age uncertainty of each level. The solid line through the data points is an additive model fitted to the observations, with prior weights given by the estimated age uncertainties. The fitted age-depth model is constrained to be monotonically decreasing with increasing depth, following the method of [@Pya2015-wr] using the **scam** package [@pya-scam-pkg]. Also shown are 25 simulations from the model posterior of the montonically constrained GAM. Each simulation from the posterior distribution of the age-model is itself a potential age-depth model which can be used to assign dates to the Small Water core. The trend model in \ref{eq:gam} can be fitted to the δ^15^N data using these new dates as a covariate, and the whole process repeated for a large number of simulations.

@@

## Further considerations

>>>>

### Example time series

To illustrate the estimation of trends in palaeoenvironmental data using GAMs, I use two proxy time series; a 150-year bulk organic matter δ^15^N time series from Small Water, and a 3000-year alkenone record from Braya-sø. Between them, the two examples, combine many of the features of interest to palaeoecologists that motivate the use of GAMs; non-linear trends and the question of when changes in the measured proxy occurred. The example analyses were all performed using the *mgcv* package and R, and the supplementary material contains a fully annotated document showing the R code used to replicate all the analyses described in the remainder of the paper.

@@

The second example time series is a 6,000 year record of alkenone unsaturation, \uk, from Braya-sø, a meromictic lake in  West Greenland [@DAndrea2011-oa]. Alkenones are long-chained unsaturated organic compounds that are produced by a small number of planktonic organisms known as haptophytes. The \uk{} unsaturation index is
\begin{equation*}
\uk{} = \frac{[C_{37:2}] - [C_{37:4}]}{[C_{37:2}] + [C_{37:3}]+ [C_{37:4}]}
\end{equation*}
where $[C_{37:x}]$ is the concentration of the alkenone with 37 carbon atoms and $x$ double carbon bonds. The relative abundance of these alkenones is known to vary with changes in water temperature, and as a result \uk{} is used as a proxy for lake- and sea-surface temperatures. For further details on the Braya-sø \uk{} record and age model see @DAndrea2011-oa. Here I use the 3,000 year \uk{} record from the PAGES 2K database [@PAGES_2k_Consortium2013-fm]. The data are presented in Figure \ref{fig:data-figure}b.

@@

In the subsequent sections, I present regression models that can be used to estimate trends in palaeoenvironmental time series and give an introduction to GAMs and how they estimate smooth, non-linear trends from time series. Next the issue of uncertainty in model-estimated trends is discussed, as is the topic of posterior simulation from a regression model. The issue of identifying periods of significant environmental change is considered and I show how this can be achieved using the first derivative of the estimated trend. Two non-standard types of spline --- adaptive smoothers and Gaussian process splines --- that are especially applicable to GAMs in the palaeoenvironmental setting are described, followed by an assessment of the the impact of age-model uncertainty on trends estimation via GAMs. Finally, I briefly discuss the application of GAM trend analysis to multivariate species abundance and compositional data.

## Regression models for palaeoenvironmental time series

A linear model for a trend in a series of $T$ observations $y_t$ at observation times $x_t$ with $t = 1, 2, \ldots, T$ is
\begin{equation} \label{eq:linear-model}
y_t = \beta_0 + \beta_1 x_t + \varepsilon_t \,,
\end{equation}
where $\beta_0$ is a constant term, the model *intercept*, representing the expected value of $y_t$ where $x_t$ is 0. $\beta_1$ is the *slope* of the best fit line through the data; it measures the rate of change in $y$ for a unit increase in $x$. The unknowns, the $\beta_j$ are commonly estimated using least squares by minimising the sum of squared errors, $\sum_t \varepsilon_t^2$. If we want to ask if the estimated trend $\beta_1$ is statistically significant, a process called *inference*, we must make further assumptions about the data (conditional upon the fitted model) or the model errors (residuals); $\varepsilon_t \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2)$. This notation indicates that the residuals $\varepsilon_t$ are *independent* and *identically distributed* Gaussian random variables with mean equal to $0$ and constant variance $\sigma^2$. In the time series setting, the assumption of independence of model residuals is often violated.

The linear model described above is quite restrictive in terms of the types of trend it can fit; essentially linear increasing or decreasing trends, or, trivially, a null trend of no change. This model can be extended to allow for non-linear trends, most notably by making $y_t$ depend on polynomials of $x_t$, for example
\begin{align} \label{eq:polynomial-model}
y_t &= \beta_0 + \beta_1 x_t + \beta_2 x_t^2 + \cdots + \beta_P x_t^P + \varepsilon_t \\
    &= \beta_0 + \sum_{p = 1}^P \beta_p x_t^p  + \varepsilon_t \,, \nonumber
\end{align}
where polynomials of $x_t$ up to order $P$ are used. This model allows for more complex trends but it remains a fully parametric model and suffers from several problems, especially the behaviour of the fitted polynomial at the start and end of the observed series.

@@

Linear models using a range of polynomials for the Small Water data set are shown in Figure \ref{fig:polynomial-example-plot}. The low-order models ($P \in \{1, 3\}$) result in very poor fit to the data. The model with $P = 5$ does a reasonable job of capturing the gross pattern in the data, but it fails to adapt quickly enough to the decrease in δ^15^N that begins ~1940 CE, and the resulting trend is quite biased as a result. The $P = 10$th-order polynomial model is well able to capture this period of rapid change, but it does so at the expense of increased complexity in the estimated trend prior to ~1940. Additionally, this model ($P = 10$) has undesirable behaviour at the ends of the series, significantly overfitting the data, a commonly observed problem in polynomial models such as these. Finally, the choice of what order of polynomial to fit is an additional choice left for the analyst to specify; choosing the optimal $P$ is not a trivial task when the data are a time series and residual autocorrelation is likely present.

@@

Can we do better than these polynomial fits? In the remainder , I hope to demonstrate that the answer to that question is emphatically "yes"! Below I describe a coherent and consistent approach to modelling palaeoenvironmental time series using generalised additive models (GAMs), which builds upon the linear regression framework.

@@

GAMs extend the generalised linear model (GLM) --- of which the linear model described above is a special case --- where some or all of the parametric terms, the $\beta_p$, are replace by smooth functions $f_j$ of the covariates $x_j$. The GAM version of the linear model \eqref{eq:linear-model} is

@@

where $\mu_t$ is the expected value (e.g.\ the mean count or the probability of occurrence) of the random variable $Y_t$ ($\mu_t \equiv \mathbb{E}(Y_t)$) of which we have observations $y_t$. $g$ is the link function, an invertible, monotonic function, such as the natural logarithm, and $g^{-1}$ is its inverse. The link function maps values from the linear predictor scale on to the scale of the response. For example, count data are strictly non-negative integer values and are commonly modelled as a Poisson GLM/GAM using the natural log link function. On the log scale, the response can take any real value between $-\infty$ and $+\infty$, and it is on this scale that model fitting actually occurs. However we need to map these unbounded values back on to the non-negative response scale the. The inverse of the log link function, the exponential function, achieves this and maps values to the interval 0--$\infty$.

In \eqref{eq:gam}, we further assume that the observations are drawn from a member of the exponential family of distributions --- such as the Poisson for count data, the binomial for presence/absence or counts from a total --- with expected value $\mu_t$ and possibly some additional parameters $\Theta$ ($y_t \sim \text{EF}(\mu_t, \Theta)$). Additionally, many software implementations of the above model also allow for distributions that are not within the exponential family but which can be fitted using an algorithm superficially similar to the one used to fit GAMs to members of the exponential family. Common examples of such extended families include the negative binomial distribution (for overdispersed counts) and the beta distribution (for true proportions or other interval-bounded data).

@@

We want the fitted trends for the Small Water δ^15^N and Braya-sø \uk{} time series to be wiggly functions, whose shape is informed by the data themselves. These wiggly functions are smooths, which we denote by $f(x_t)$. We would like to represent the smooths in a way that \eqref{eq:gam} is a linear model so that we can estimate the model within the well-studied GLM framework, which is achieved by using a *basis*. A basis is a set of functions that collectively span a space of functions that, we hope, contains the true $f(x_t)$ or a close approximation to it. The functions in the basis are known as *basis functions*, and arise from a *basis expansion* of a covariate. Writing $b_j(x_t)$ as the $j$th basis function of $x_t$, the smooth $f(x_t)$ can be represented as a weighted sum of basis functions

$$
f(x_t) = \sum_{j = 1}^{k} b_j(x_t) \beta_j \,,
$$

@@

The polynomial model we encountered earlier is an example of a statistical model that uses a basis expansion. For the cubic polynomial ($P = 3$) fit shown in Figure \ref{fig:polynomial-example-plot} there are in fact 4 basis functions: $b_1(x_t) = x_t^0 = 1$, $b_2(x_t) = x_t$, $b_3(x_t) = x^2_t$, and $b_4(x_t) = x_t^3$. Note that here $b_1(x_t)$ is constant and is in fact the model intercept, $\beta_0$, in the linear model \eqref{eq:polynomial-model}, and the weights are the estimated coefficients in the model, the $\beta_j$.

As we have already seen, polynomial basis expansions do not necessarily lead to well-fitting models unless the true function $f$ is itself a polynomial. One of the primary criticisms is that polynomial basis functions are global; the value of $f$ at time point $x_t$ affects the value of $f$ at time point $x_{t+s}$ even if the two time points are at opposite ends of the series. There are many other bases we could use; here I discuss one such set of bases, that of splines.

@@

The CRS can be parameterized in a number of different ways. One requires a knot at each unique data value in $x_t$, which is computationally inefficient. Another way of specifying a CRS basis is to parameterize in terms of the value of the spline at the knots. Typically in this parametrization there are many fewer knots than unique data, with the knots distributed evenly over the range of $x_t$ or at the quantiles of $x_t$. Placing knots at the quantiles of $x_t$ has the effect of placing a greater number of knots where the data is densest.

@@

Cubic regression splines, as well as many other types of spline, require the analyst to choose the number and location of the knots that parametrise the basis. Thin plate regression splines (TPRS) remove this element of subjectivity when fitting GAMs. Thin plate splines were introduced by @Duchon1977-jr and, as well as solving the knot selection problem, have several additional attractive properties in terms of optimality and their ability to estimate a smooth function of two or more variables, leading to smooth interactions between covariates. However, thin plate splines have one key disadvantage over CRS; thin plate splines have as many unknown parameters as there are unique combinations of covariate values in a data set. It is unlikely in the extreme that any real data problem would involve functions of such complexity that they require as many basis functions as data. It is much more likely that the true functions that we attempt to estimate are far simpler than the set of functions representable by 1 basis function per unique data value. From a practical point of view, it is also highly inefficient to carry around all these basis functions whilst model fitting, and the available computational resources would become quickly exhausted for large time series with many observations.

@@

The truncation suggested by @Wood2003-qy is not without cost; the eigen-decomposition and related steps can be relatively costly to set up for large data sets. For data sets of similar size to the two examples used here, the additional computational effort required to set up the TPRS basis over the CRS basis will not be noticeable. For highly resolved series containing more than 500--1000 observations the truncation may be too costly computationally. In such instances, little is lost by moving to the CRS basis, with the same number of knots as the rank of the desired thin plate spline, either distributed evenly throughout the data or at the quantiles of the data, with the benefit of considerably reduced set up time for the basis.

To fit a GAM using either of the two regression spline bases described above, the analyst is generally only required to the specify the size (rank) of the basis expansion required to represent or closely approximate the true function $f$. With practice and some knowledge of the system from which the observations arise, it can be relatively easy to put an upper limit on the expected complexity of the true trend in the data. Additionally, the number available data points places a constraint on the upper limit of the size of basis expansion that can be used.

In practice, the size of the basis is an upper limit on the expected complexity of the trend, and a simple test is available to check if the basis used was sufficiently large [@Pya2016-rk]. This test is available via the `gam.check()` function in *mgcv* for example, and essentially looks at whether there is any additional nonlinearity or structure in the residuals that can be explained by $x_t$. Should a smooth term in the fitted model fail this test the model can be refitted using a larger basis expansion, say by doubling the value of `k` (the rank) used to fit the original. Note also that a smooth might fail this test whilst using fewer effective degrees of freedom than the maximum possible for the dimension of basis used. This may happen when the true function lies at the upper limit of the set of functions encompassed by the size of basis used. Additionally, a basis of size $2k$ encompasses a richer space of functions of a given complexity than a basis of size $k$ [@Wood2017-qi]; increasing the basis dimension used to fit the model may unlock this additional function space resulting in a better fitting model whilst using a similar number of effective degrees of freedom.

@@

Having identified regression splines as a useful way to represent $f$, we next need a way to decide how wiggly the fitted trend should be. A backwards elimination approach to sequentially remove knots or basis functions might seem appropriate, however such an approach would likely fail as the resulting sequence of models would not be strictly nested, precluding many forms of statistical comparison. Alternatively, we could keep the basis dimension at a fixed size but guard against fitting very complex models through the use of a wiggliness penalty.

@@

$$\mathcal{L}_p(\boldsymbol{\beta}) = \mathcal{L}(\boldsymbol{\beta}) - \frac{1}{2} \lambda\boldsymbol{\beta}^{\mathsf{T}}\mathbf{S}\boldsymbol{\beta}\, .
$$

@@

To fully automate smoothness selection for $f(x_t)$ we need to estimate $\lambda$. There are two main ways that $\lambda$ can be automatically chosen during model fitting. The first way is to choose $\lambda$ such that it minimises the prediction error of the model. This can be achieved by choosing $\lambda$ to minimise Akaike's information criterion (AIC) or via cross-validation (CV) or generalized cross-validation (GCV). GCV avoids the computational overhead inherent to CV of having to repeatedly refit the model with one or more observations left out as a test set. Minimising the GCV score will, with a sufficiently large data set, find a model with the minimal prediction error [@Wood2017-qi]. The second approach is to treat the smooth as a random effect, in which $\lambda$ is now a variance parameter to be estimated using maximum likelihood (ML) or restricted maximum likelihood (REML) [@Wood2011-kn; @Wood2016-fx].

@@

## Fitting GAMs (Need a better title)

@@

Note that now I have relaxed the i.i.d. assumption and have introduced $\boldsymbol{\Lambda}$, a correlation matrix that is used to model autocorrelation in the residuals. The δ^15^N values are irregularly spaced in time and as a result a correlation structure that can handle the uneven spacing of the samples is needed. A continuous time first-order autoregressive process (CAR(1)) is a reasonable choice; it is the continuous-time equivalent of the first-order autoregressive process (AR(1)) and, simply stated, models the correlation between any two residuals as and exponentially decreasing function of $h$ ($\phi^h$), where $h$ is the amount of separation in time between the residuals. $h$ may be a real valued number in the CAR(1), which is how it can accommodate the irregular separation of samples in time. $\phi$ controls how quickly the correlation between any two residuals declines as a function of their separation in time and is an additional parameter that will be estimated during model fitting. The model in \eqref{eq:small-gam} was fitted using the `gamm()` function [@Wood2004-zv] in the *mgcv* package [@Wood2017-qi] for R [@R-core-team].

@@

plot_grid(small_fitted, braya_fitted, ncol = 1, labels = "auto", align = "hv", axis = "lr")
```

The fitted trend is shown in Figure \ref{fig:plot-fitted-models}a, and well-captures the strong pattern in the data. The trend is statistically significant (estimated degrees of freedom = `r round(smallSumm[["s.table"]][1,2], 3)`; $F$ = `r round(smallSumm[["s.table"]][1,3], 3)`, approximate $p$ value = $\ll$ 0.0001). However further analysis of the fitted model is required to answer the other questions posed earlier about the timing of change and whether features in the trend can be distinguished from random noise. I discuss these issues shortly.

@@

The \uk{} data present a more difficult data analysis challenge than the δ^15^N time series because of the much more complex variation present. Fitting the same model as the Small Water example, \eqref{eq:small-gam}, to the \uk{} data resulted in the unsatisfactory fit shown as the very smooth line in Figure \ref{fig:plot-fitted-models}b (labelled GAMM (CAR(1))). Further problems were evident with this model fit --- the covariance matrix of the model was non-positive definite, a sure sign of problems with the fitted model. Refitting with a smaller basis dimension (*k* = 20) for the trend term resulted in a model with a positive-definite covariance matrix for the model variance-covariance terms, but the estimated value of of the CAR(1) parameter $\phi$ = `r round(brayaPhi[2], 2)` was exceedingly uncertain (95% confidence interval `r round(brayaPhi[1], 2)` -- `r round(brayaPhi[3], 2)`!)

@@

To investigate the first issue the GCV and REML scores for an increasing sequence of values of the smoothness penalty $\lambda$ were evaluated for the standard GAM (equation \eqref{eq:gam}) fit to the \uk{} time series. The resulting profiles are shown in Figure \ref{fig:trace-smoothness-parameters}, with the selected value of the penalty shown by the vertical line. The GCV score profile suggests that the potential for under smoothing identified by @Reiss2009-fk is unlikely to apply here. Addressing the violation of the independence assumption is more difficult; the standard errors of the model coefficients are most likely anti-conservative, which presents a significant problem if we wish to determine which of the wiggles in the fitted \uk{} trend can be identified as real.

@@

While there is not much we can do within the GAM framework to model a series that contains both smooth trends and step-like responses, an adaptive smoother can help address problems where the time series consists of periods of very rapid change in the mean combined with periods of complacency or relatively little change. As suggested by their name, adaptive smoothers can adjust to changes in the wiggliness of the time series. This adaptive behaviour is achieved by making the smoothness parameter $\lambda$ itself depend smoothly on $x_t$; in other words, the adaptive smoother allows the wiggliness of the estimated trend to vary smoothly over time. This is achieved by using multiple penalties and smoothness parameters, which have the effect of letting the strength of the wiggliness vary smoothly over time. Whilst this does allow the estimated trend to adapt to periods of rapid change in the response, adaptive smoothers make significant demands on the data; if we used $m$ smoothness penalties to allow the wiggliness to vary over a time series, it would be like estimating $m$ separate smooths from chunks of the original series each of length $n/m$. In a practical sense, this limits the use of adaptive splines in palaeoecology to proxies that are readily enumerated, such as the biogeochemical proxies used in the two example data sets.

@@

Figure \ref{fig:small-scam-fit-plots}a shows the estimated dates (in Years CE) for `r nrow(swAge)` levels in the Small Water core dated using ^210^Pb. The vertical bars show the estimated age uncertainty of each level. The solid line through the data points is an additive model fitted to the observations, with prior weights given by the estimated age uncertainties. The fitted age-depth model is constrained to be monotonically decreasing with increasing depth, following the method of [@Pya2015-wr] using the *scam* package [@pya-scam-pkg]. Also shown are 25 simulations from the model posterior of the montonically constrained GAM. Each simulation from the posterior distribution of the age-model is itself a potential age-depth model which can be used to assign dates to the Small Water core. The trend model in \eqref{eq:gam} can be fitted to the δ^15^N data using these new dates as a covariate, and the whole process repeated for a large number of simulations.

@@
